---
title: "Supplementary Material for Two Stage Adaptive Design for Prognostic Biomarker Signature with Survival Endpoint"
author: "Biyue Dai"
date: "July 22, 2019"
output: html_document
---

This article provides a guidance of the supplementary R functions for the manuscript "Two-Stage Adaptive Design for Prognostic Biomarker Signatures with a Survival Endpoint" 

# **Documentation of Functions**

> **two_stage_permutation()**

#### Description

Analyze a given data set using the two-stage design. Inference in the first stage is conducted via permutation.

#### Arguments

 **survY**: output from surv function in survival package

 **X**: a matrix or data frame of variables of interest, without missing value

 **itrB**: number of iteration used in permutation

 **seed**: seed number for reproducing results (default: NULL)

 **nfold**: number of folds used for cross-validation (default: 10)

 **S**: proportion of samples used in first stage (default: 0.5)

 **alpha_1**: level of test in first stage (default: 0.25)

 **alpha_2**: level of test in second stage (default: 0.2)
 
 **printmodel**: logical value. TRUE indicates the final signature will be print out (default: FALSE)
  
> **two_stage_bootstrap()**

#### Description

Analyze a given data set using the two-stage design. Inference in the first stage is conducted via bootstrap.

#### Arguments

 **survY**: output from surv function in survival package

 **X**: a matrix or data frame of variables of interest, without missing value

 **itrB**: number of iteration used in permutation

 **seed**: seed number for reproducing results (default: NULL)

 **nfold**: number of folds used for cross-validation (default: 10)

 **S**: proportion of samples used in first stage (default: 0.5)

 **alpha_1**: level of test in first stage (default: 0.25)

 **alpha_2**: level of test in second stage (default: 0.2)
 
 **null_C **: The value of C Index under the null hypothesis of interest (default: 0.5)
 
 **printmodel **: Logical value that allows the user to print out the model that is built at the end of first stage. (default: FALSE)

> **.cvFolds()**
  
  Lower level function that assigns fold IDs given number of folds.

> **cvcox()**
  
  Lower level function that calculates cross-validated linear predictors from Cox Regression

# **Examples**

```{r, warning= FALSE, message= FALSE}
# load functions
source("functions.R")
# load data
data(lung) # built-in data set in survival package

lung <- na.omit(lung) # for simplicity, I only took the complete cases
survY <- Surv(lung$time, lung$status)
X <- lung[,c("age","sex","ph.ecog","ph.karno","meal.cal","wt.loss")]
```

Use Permuation Test in the First Stage:

```{r, warning= FALSE, message= FALSE}
two_stage_permutation(survY, X, itrB = 100, S =0.5, seed = 666, printmodel = TRUE)
```


Use Bootstrap in the First Stage, testing Null Hypothesis: H0: C = 0.5
```{r, warning= FALSE, message= FALSE}
two_stage_bootstrap(survY, X, itrB = 100, seed = 666, S =0.5)
```

Use Bootstrap in the First Stage, testing Null Hypothesis: H0: C = 0.6
```{r, warning= FALSE, message= FALSE}
two_stage_bootstrap(survY, X, itrB = 100, seed = 666, S =0.5, null_C = 0.6)
```

The output contains the following values:

survC: cross-validated C Index of the model built in Stage 1

threshold: the rejection threshold in permutation and bootstrap distribution 

early_stop: indicator of whether the trial is stopped early or not. 1: stopped early; 2: continue to second stage

C_2: C Index in stage 2. It would be NA if the study is stopped early.

se_2: standard error of C Index in stage 2. It would be NA if the study is stopped early.

final_reject: Indicator of whether the signature is finally validated. 1: validated; 2: validation has failed.

cens, cens1, cens2: censoring proportion overall, for stage 1 and for stage 2.

If **printmodel** is set to be true, then the final model that is built based on coxph will be printed.