---
title: "Supplementary Material II: Simulation"
subtitle: "Two Stage Design for Prognostic Biomarker Signatures with Survival Endpoint"
author: "Biyue Dai"
date: "August 21, 2019"
output: html_document
---

This article provides a guidance of the supplementary R functions for the manuscript "Two-Stage Adaptive Design for Prognostic Biomarker Signatures with a Survival Endpoint" 

# **Documentation of Functions**

> **two_stage_sim()**

#### Description

Conduct simulation for the proposed two-stage design for given scenarios.

#### Arguments

**nn**: Total sample size 

**rate_l**: rate parameter in the Exponential Distribution that generates the Time of Event Occurs (default = 0.025)

**rate_c**: rate parameter in the Exponential Distribution that generates the Time of Censoring (default = 0.025)

**beta**: the true regression coefficients

**mean**: parameter in generating the dependent variables (default = 0)

**sd**: parameter in generating the dependent variables (default = log(1.5))

**itrB**: number of iterations used for bootstrap or permutation resampling (default = 200)

**nfold**: number of folds for cross-validation (default = 10)

**S**: proportion of samples used in first stage (default: 0.5)

**alpha_1**: level of test in first stage (default: 0.25)

**alpha_2**: level of test in second stage (default: 0.2)
 
**null_C**: The value of C Index under the null hypothesis of interest for bootstrap (default: 0.5)

**method**: character that takes value of "permutation" or "bootstrap".

**rep = 500**: total number of iterations for the Monte Carlo Simulation

**seedID**: seed number for reproducing results (default: NULL)

> **two_stage_permutation()**

#### Description

Analyze a given data set using the two-stage design. Inference in the first stage is conducted via permutation.

#### Arguments

 **survY**: output from surv function in survival package

 **X**: a matrix or data frame of variables of interest, without missing value

 **itrB**: number of iteration used in permutation

 **seed**: seed number for reproducing results (default: NULL)

 **nfold**: number of folds used for cross-validation (default: 10)

 **S**: proportion of samples used in first stage (default: 0.5)

 **alpha_1**: level of test in first stage (default: 0.25)

 **alpha_2**: level of test in second stage (default: 0.2)
 
 **printmodel**: logical value. TRUE indicates the final signature will be print out (default: FALSE)
  
> **two_stage_bootstrap()**

#### Description

Analyze a given data set using the two-stage design. Inference in the first stage is conducted via bootstrap.

#### Arguments

 **survY**: output from surv function in survival package

 **X**: a matrix or data frame of variables of interest, without missing value

 **itrB**: number of iteration used in permutation

 **seed**: seed number for reproducing results (default: NULL)

 **nfold**: number of folds used for cross-validation (default: 10)

 **S**: proportion of samples used in first stage (default: 0.5)

 **alpha_1**: level of test in first stage (default: 0.25)

 **alpha_2**: level of test in second stage (default: 0.2)
 
 **null_C **: The value of C Index under the null hypothesis of interest (default: 0.5)

> **tuning_C()**

#### Description

Function that helps tune the beta coefficients to hit the target level of C for the alternative hypothesis. For a given vector of beta, only the last coefficient will be tuned and all other coefficients will be fixed.

#### Arguments

**nn**: number of samples used as the population (default = 10000)

**mean**: parameter in generating the dependent variables (default = 0)

**sd**: parameter in generating the dependent variables (default = log(1.5))

**rate_l**: rate parameter in the Exponential Distribution that generates the Time of Event Occurs (default = 0.025)

**p**: number of covariates
                     
**initial_beta**: a vector of beta coefficients to start with (for p and initial_beta, only of the two arguments is needed)

**targetC**: target value of C Index of interest
                      
**by, start, end**: parameters that create a sequence of candidate values for the last coefficient. start: where the sequence starts; end: where the sequence ends; by: the increments for the sequence.
                    
**seedID**: seed number for reproducing results (default: NULL)


> **tuning_censoring()**

Function that helps tune the distribution of censoring time so that censoring hits the targete proportion.

#### Arguments

**nn**: number of samples used as the population (default = 1000)

**mean**: parameter in generating the dependent variables (default = 0)

**sd**: parameter in generating the dependent variables (default = log(1.5))

**rate_l**: rate parameter in the Exponential Distribution that generates the Time of Event Occurs (default = 0.025)

**by, start, end**: parameters for creating the sequence that can be used to tune the last coefficient. start: where the sequence starts; end: where the sequence ends; by: the increments for the sequence.

**p**: number of covariates
                     
**beta**: a vector of beta coefficients

**target**: target value of censoring proportion
                      
**seed**: seed number for reproducing results (default: NULL)


> **.cvFolds()**
  
  Lower level function that assigns fold IDs given number of folds.

> **cvcox()**
  
  Lower level function that calculates cross-validated linear predictors from Cox Regression

# **Examples**

### Example 2: Simulation Under the Null Hypothesis: C = 0.5

We first illustrate how to use the permutation version

```{r, warning= FALSE, message= FALSE}
source("functions.R")
```

```{r}
result1 <- two_stage_sim(nn = 300, # sample size
              rate_l = 0.025, rate_c = 0.025,
              beta = rep(0,10),mean = 0, sd = log(1.5),
              nfold = 10,
              itrB = 50, #recommend >= 200
              S = 0.5,
              alpha_1 = 0.25, alpha_2 = 0.2,
              null_C = 0.5,
              method = "permutation",
              rep = 10, #recommend >= 500
              seedID = 999
)
head(result1)
```

As a demonstration, we only specified 5 iterations for the Monte Carlo simulation. Usually it takes >= 500 iterations to evaluate operation characteristics.

To compute type one error rate:
```{r}
mean(result1[,"final_validation"])
```
To compute probability of early stopping:
```{r}
mean(result1[,"early_stop"])
```

### Example 3: Simulate Samples with the population C Index > 0.5

Generating data with C > 0.5 is less straignthfoward than generating data with C = 0.5. The difficulty comes from hitting the right level of the population C Index and also maintaining the correct proportion of censoring.

We provided two functions \textbf{tuning_C} and \textbf{tuning_censoring} to select the parameters that can generate the targeted level of C.

#### Step 1: Tune the Regression coefficients with the \textbf{tuning_C} function so that the Population C Index reaches a targeted value.

In this example, we aim at having the C Index to be **0.6**. We want to have **5** covariates in the model.

```{r}
tuned_C <- tuning_C(nn = 10000,
                    mean = 0,
                    sd = log(1.5),
                    rate_l = 0.025,
                    p = 5,
                    by = 0.01,
                    start = 0.5,
                    end = 1.1,
                    targetC = 0.6,
                    seed = 123)
tuned_C
```

The first two outputs yield the values of the beta coefficients that should be used later:

```{r}
tuned_C$beta
tuned_C$beta_j
```

#### Step 2: Tune Censoring Distribution with the \textbf{tuned_censor} function to Hit the Targeted Censoring Percentage

Given the beta coefficients produced by \textbf{tuned_C}, We need to tune the rate parameter in the censoring distribution so that 30 percent of the data is censored. 

```{r}
tuned_censor<- tuning_censoring(nn = 10000,
                                mean = 0,
                                sd = log(1.5), 
                                rate_l = 0.025,
                                start = 0.001,
                                end = 0.025*0.3*(1-0.3) + 0.05,
                                by = 0.0001,
                                beta = c(tuned_C$beta,tuned_C$beta_j),
                                target = 0.3,
                                seed = 111)
tuned_censor
```

Here we take rate_c from the output as the rate parameter for the censoring distribution.

```{r}
tuned_censor$rate_c
```


#### Step 3: Apply the tuned parameters in simulation:

We can use these data to conduct a simulation with the permutation version of the design. In this scenario, the null Hypothesis is C = 0.5. Since the data is generated under C = 0.6, we will be looking at the power of the study.

```{r, echo = FALSE}
load("H:\\Courses\\TACO\\simulation_code\\results.RData")
```

```{r, eval = FALSE}
result_P <- two_stage_sim(nn = 500, # sample size
              rate_l = 0.025, rate_c = tuned_censor$rate_c,
              beta = c(tuned_C$beta,tuned_C$beta_j),mean = 0, sd = log(1.5),
              nfold = 10,
              itrB = 200, #recommend >= 200
              S = 0.5,
              alpha_1 = 0.25, alpha_2 = 0.2,
              method = "permutation",
              rep = 500, #recommend >= 500
              seedID = 999
              )
head(result_P)
```

```{r, echo = FALSE}
head(result_P)
```

To compute Power:
```{r}
mean(result_P[,"final_validation"])
```

To compute probability of Early Stopping:
```{r}
mean(result_P[,"early_stop"])
```

We can use these data to conduct a simulation with the bootstrap version. In this scenario, we can set the null Hypothesis to be C = 0.6. Since the data is generated under C = 0.6, we will be looking at the type one error rate of the study.

```{r, eval = FALSE}
result_B <- two_stage_sim(nn = 500, # sample size
              rate_l = 0.025, rate_c =  tuned_censor$rate_c,
              beta = c(tuned_C$beta,tuned_C$beta_j),mean = 0, sd = log(1.5),
              nfold = 10,
              itrB = 200, #recommend >= 200
              S = 0.5,
              alpha_1 = 0.25, alpha_2 = 0.2,
              null_C = 0.6,
              method = "bootstrap",
              rep = 500, #recommend >= 500
              seedID = 123
              )
head(result_B)
```

```{r, echo = FALSE}
head(result_B)
```

To compute Type one error rate:
```{r}
mean(result_B[,"final_validation"])
```

To compute probability of Early Stopping:
```{r}
mean(result_B[,"early_stop"])
```
